---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `{rpt}` ([R]{.underline} [p]{.underline}ackage [t]{.underline}emplate)

<!-- badges: start -->

```{r, echo = FALSE, message = FALSE, results='asis'}
library(badger)
cat(
  badge_lifecycle(stage = "experimental", alt = "Lifecycle: experimental"),
  badge_github_actions(alt = "R-CMD-check", action = "R-CMD-check.yaml"),
  badge_codecov(alt = "Codecov test coverage", branch = "main"),
  badge_codefactor(alt = "CodeFactor", ref = "ucd-serg/rpt"), # case matters
  badge_cran_release(alt = "CRAN status"),
  badge_cran_download(type = "grand-total"),
  badge_cran_download(type = "last-month"),
  badge_cran_download(type = "last-week"),
  badger::badge_license()
)
```

<!-- badges: end -->

The goal of `{rpt}` is to ...

## Installation

You can install the development version of `{rpt}` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("UCD-SERG/rpt")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(rpt)
## basic example code
```

## Development

### Building the Documentation Site

This package uses [altdoc](https://altdoc.etiennebacher.com/) with [Quarto](https://quarto.org/) to build its documentation site. To build and preview the documentation locally:

```r
# Load the package
pkgload::load_all()

# Render the documentation
altdoc::render_docs()

# Preview the site
altdoc::preview_docs()
```

The documentation is automatically built and deployed to GitHub Pages via GitHub Actions when changes are pushed to the main branch.

## Other R Package Template Options

If you're looking for alternative R package templates, you may also want to consider:

- [r.pkg.template](https://github.com/insightsengineering/r.pkg.template/) - A comprehensive R package template from Insights Engineering

## Code of Conduct

Please note that the `{rpt}` project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
