---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `{ettbc}`

<!-- badges: start -->

```{r, echo = FALSE, message = FALSE, results='asis'}
library(badger)
cat(
  badge_lifecycle(stage = "experimental", alt = "Lifecycle: experimental"),
  badge_github_actions(alt = "R-CMD-check", action = "R-CMD-check.yaml"),
  badge_codecov(alt = "Codecov test coverage", branch = "main"),
  badge_codefactor(alt = "CodeFactor", ref = "ucdavis/ettbc"), # case matters
  badger::badge_license()
)
```

<!-- badges: end -->

`{ettbc}` emulates a target trial of screening mammography for breast cancer,
using the **clone-censor-reweight** methodology of
[García-Albéniz et al. (2020)](https://doi.org/10.7326/M18-1199). That study
estimated the effect of continuing (vs. stopping) annual screening mammography
on breast cancer mortality in Medicare beneficiaries older than 70.

The package provides the analytic building blocks of that pipeline:

- `clone_censor()` --- clone each participant into the STOPBASE and CONTINUE
  arms and apply each arm's censoring rule.
- `expand_to_long()` --- expand the cloned data to one row per
  participant-arm-month for discrete-time survival analysis.
- `compute_ipw_weights()` --- stabilized inverse-probability-of-censoring
  weights.
- `fit_outcome_hr()` --- IPW-weighted pooled logistic regression for the
  arm hazard ratio.
- `predict_survival_unadjusted()`, `predict_survival_baseline_adjusted()`,
  `predict_survival_ipw()` --- g-computation marginal survival curves.
- `bootstrap_ci()` --- percentile bootstrap confidence intervals.
- `false_positives()` --- false-positive rates for histological evaluations.
- `extract_screening_mammograms()` and companions --- templates for
  extracting mammogram events from Medicare claims by HCPCS code.

The package is experimental and not yet an end-to-end reproduction of the
published analysis; see the open issues for the remaining gaps.

## Installation

You can install the development version of `{ettbc}` from
[GitHub](https://github.com/ucdavis/ettbc) with:

``` r
# install.packages("pak")
pak::pak("ucdavis/ettbc")
```

## Example

`{ettbc}` ships three synthetic datasets (`cohort`, `screening_mammograms`,
`diagnostic_mammograms`) that mimic the structure of the original study data.
The clone-censor step turns the one-row-per-participant cohort into two rows
per participant --- one for each trial arm:

```{r example}
library(ettbc)

cloned <- clone_censor(
  cohort,
  screening_mammograms,
  diagnostic_mammograms
)

nrow(cloned) # two rows per participant
head(cloned[, c("id", "arm", "start_month", "end_month", "fup", "bc_died")])
```

`expand_to_long()` then prepares the data for the inverse-probability-weighted
survival analysis:

```{r example-long}
long <- expand_to_long(cloned)
head(long[, c("id", "arm", "month", "month2", "dead_t1")])
```

See the article
["Using ettbc"](https://ucdavis.github.io/ettbc/articles/using-ettbc.html)
for the full walkthrough.

## Development

### Building the Documentation Site

This package uses [altdoc](https://altdoc.etiennebacher.com/) with
[Quarto](https://quarto.org/) to build its documentation site. To build and
preview the documentation locally:

```r
# Load the package
pkgload::load_all()

# Render the documentation
altdoc::render_docs()

# Preview the site
altdoc::preview_docs()
```

The documentation is automatically built and deployed to GitHub Pages via
GitHub Actions when changes are pushed to the main branch.

## Code of Conduct

Please note that the `{ettbc}` project is released with a [Contributor Code of Conduct](https://contributor-covenant.org/version/2/1/CODE_OF_CONDUCT.html). By contributing to this project, you agree to abide by its terms.
