---
title: "Getting Started"
format:
  html: default
  revealjs:
    output-file: getting-started-revealjs.html
  docx: default
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ettbc)
```

## Introduction

`{ettbc}` emulates a target trial of screening mammography for breast cancer,
following the clone-censor-reweight methodology of García-Albéniz et al.
(2020). This vignette walks through the first two steps of the pipeline on
the package's synthetic example data. For the full analysis --- including
inverse-probability weighting, the outcome model, survival curves, and
bootstrap confidence intervals --- see the article "Using ettbc".

## Example Data

The package ships three synthetic datasets that mimic the structure of the
original study data:

```{r}
# One row per participant
head(cohort)
```

## Step 1: Clone and Censor

`clone_censor()` replicates each participant into two trial arms --- STOPBASE
(stop screening at entry) and CONTINUE (continue annual screening) --- and
applies each arm's censoring rule:

```{r}
cloned <- clone_censor(
  cohort,
  screening_mammograms,
  diagnostic_mammograms
)

nrow(cloned) # two rows per participant
head(cloned[, c("id", "arm", "start_month", "end_month", "fup", "bc_died")])
```

## Step 2: Expand to Long Format

`expand_to_long()` converts the cloned data to one row per
participant-arm-month, ready for discrete-time survival analysis:

```{r}
long <- expand_to_long(cloned)
head(long[, c("id", "arm", "month", "month2", "dead_t1")])
```

## Next Steps

From the long-format data you can compute inverse-probability weights with
`compute_ipw_weights()`, fit the outcome model with `fit_outcome_hr()`,
estimate marginal survival curves with the `predict_survival_*()` functions,
and obtain confidence intervals with `bootstrap_ci()`. The article
"Using ettbc" demonstrates the full pipeline.
