---
title: "Quarto Vignette Example"
format:
  html: default
  revealjs:
    output-file: quarto_vignette-revealjs.html
  docx: default
vignette: >
  %\VignetteIndexEntry{Quarto Vignette Example}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(rpt)
```

## Introduction

This vignette demonstrates how to use Quarto (`.qmd`) files as vignettes in R packages. Quarto is the next-generation version of R Markdown, offering enhanced features and better integration with multiple programming languages.

## Why Use Quarto?

Quarto offers several advantages over traditional R Markdown:

1. **Enhanced features**: Better support for cross-references, callouts, and other advanced formatting
2. **Multiple engines**: Works with R, Python, Julia, and Observable
3. **Modern syntax**: Cleaner YAML and chunk options using `#|` syntax
4. **Better publishing**: Improved website and book publishing capabilities

{{< slidebreak >}}

## Basic Example

Here's a basic example using the package:

```{r example}
# Create some sample data
sample_data <- c(1, 2, 3, 4, 5, 10, 15, 20)

# Use the example function
result <- example_function(sample_data)
result
```

## Quarto Features

### Callouts

::: {.callout-note}
This is a callout block that highlights important information.
:::

::: {.callout-tip}
Quarto supports multiple types of callouts: note, tip, warning, important, and caution.
:::

{{< slidebreak >}}

### Code Chunks with Modern Syntax

You can use the modern `#|` syntax for chunk options:

```{r}
#| label: fig-plot
#| fig-cap: "Example plot showing data distribution"
#| echo: true

# Create a simple plot
plot(sample_data, 
     main = "Sample Data", 
     xlab = "Index", 
     ylab = "Value",
     pch = 19,
     col = "steelblue")
```

## Equations

Quarto has excellent support for mathematical equations using LaTeX syntax:

Inline equation: $E = mc^2$

Display equation:

$$
\bar{x} = \frac{1}{n}\sum_{i=1}^{n} x_i
$$

## Tables

| Feature | R Markdown | Quarto |
|---------|-----------|--------|
| YAML | Standard | Enhanced |
| Chunks | `{r, option=value}` | `#| option: value` |
| Publishing | Good | Excellent |
| Cross-refs | Limited | Native |

: Comparison of R Markdown and Quarto {#tbl-comparison}

## Summary

Quarto vignettes provide a modern, feature-rich way to document R packages. They integrate seamlessly with documentation systems and offer enhanced capabilities for technical documentation.

## References

For more information, see:

- [Quarto Documentation](https://quarto.org/)
- [Using Quarto with R](https://quarto.org/docs/computations/r.html)
- @R-base
- @R-rpt
