---
title: "Advanced Quarto Article"
author: "UC Davis Seroepidemiology Research Group (UCD-SERG)"
date: "`r Sys.Date()`"
format:
  revealjs:
    output-file: quarto_article-revealjs.html
  html: default
  docx: default
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

```{r setup}
#| message: false
library(rpt)
```

## Overview

This article demonstrates advanced features of Quarto for package documentation. Unlike vignettes, articles are only available on the documentation website and are not included in the package bundle.

## When to Use Articles vs Vignettes

**Vignettes** should be used for:

- Core package documentation
- Essential usage examples
- Content that users need offline

**Articles** are better for:

- Extended tutorials
- Case studies and detailed examples
- Content with large datasets or external dependencies
- Supplementary material

## Advanced Quarto Features

### Cross-References

Quarto makes it easy to create cross-references. For example, see @fig-scatter for a scatter plot and @tbl-summary for a summary table.

```{r}
#| label: fig-scatter
#| fig-cap: "Scatter plot demonstrating cross-referencing"
#| echo: true

# Generate sample data
set.seed(123)
x <- rnorm(100)
y <- 2 * x + rnorm(100, 0, 0.5)

# Create scatter plot
plot(x, y, 
     main = "Example Scatter Plot",
     xlab = "X Variable",
     ylab = "Y Variable",
     pch = 16,
     col = rgb(0.2, 0.4, 0.6, 0.6))
abline(lm(y ~ x), col = "red", lwd = 2)
```

### Summary Tables

```{r}
#| label: tbl-summary
#| tbl-cap: "Summary statistics for example data"

data_summary <- data.frame(
  Statistic = c("Mean", "Median", "SD", "Min", "Max"),
  X = c(mean(x), median(x), sd(x), min(x), max(x)),
  Y = c(mean(y), median(y), sd(y), min(y), max(y))
)

knitr::kable(data_summary, digits = 3)
```

As shown in @tbl-summary, the variables have similar distributions.

### Code Folding

You can make code chunks foldable:

```{r}
#| code-fold: true
#| code-summary: "Show the code for data preparation"

# Prepare more complex data
complex_data <- data.frame(
  id = 1:50,
  group = rep(c("A", "B"), each = 25),
  value1 = rnorm(50, 100, 15),
  value2 = rnorm(50, 50, 10)
)

# Calculate group statistics
group_stats <- aggregate(
  cbind(value1, value2) ~ group, 
  data = complex_data, 
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)

print(group_stats)
```

### Tabsets

::: {.panel-tabset}

#### Plot View

```{r}
#| echo: false
boxplot(value1 ~ group, data = complex_data,
        main = "Value 1 by Group",
        xlab = "Group",
        ylab = "Value 1",
        col = c("lightblue", "lightgreen"))
```

#### Summary View

```{r}
#| echo: false
summary(complex_data)
```

#### Raw Data

```{r}
#| echo: false
head(complex_data, 10)
```

:::

### Advanced Callouts

::: {.callout-important}
## Important Considerations

When using this package in production:

1. Always validate input data
2. Check for missing values
3. Consider computational complexity
:::

::: {.callout-warning}
## Performance Warning

Large datasets may require additional memory and processing time.
:::

### Columns Layout

:::: {.columns}

::: {.column width="50%"}
**Left Column**

This demonstrates a two-column layout in Quarto.

- Feature 1
- Feature 2
- Feature 3
:::

::: {.column width="50%"}
**Right Column**

You can place different content in each column.

```{r}
#| echo: false
hist(rnorm(100), 
     main = "Distribution",
     col = "lightblue",
     border = "white")
```
:::

::::

## Working with Package Functions

Here's how to use the package's example function with different inputs:

```{r}
# Numeric vector
example_function(c(5, 10, 15, 20, 25))

# Using with generated data
random_data <- runif(10, min = 0, max = 100)
example_function(random_data)
```

## Code Annotations

Quarto supports code annotations for detailed explanations:

```{r}
# <1>
data <- c(1, 2, 3, 4, 5) # <2>
result <- example_function(data) # <3>
print(result) # <4>
```
1. Load sample data
2. Create a simple numeric vector
3. Apply the example function from the package
4. Display the results

## Mathematical Notation

Quarto excels at mathematical notation. Here's an example of a statistical formula:

The standard error of the mean is calculated as:

$$
SE = \frac{s}{\sqrt{n}}
$$

where $s$ is the sample standard deviation and $n$ is the sample size.

For a confidence interval:

$$
CI = \bar{x} \pm t_{\alpha/2, n-1} \cdot SE
$$

## Tips and Best Practices

::: {.callout-tip collapse="true"}
## Expanding tip section

Click to expand this section for additional tips:

1. Use descriptive chunk labels for better organization
2. Set global chunk options at the beginning
3. Use cross-references to link related content
4. Leverage callouts for important information
5. Consider code folding for long code blocks
:::

## Conclusion

This article demonstrates the advanced capabilities of Quarto for creating rich, interactive documentation for R packages.

## Learn More

- Visit the [package website](https://ucd-serg.github.io/rpt/)
- Check out the [GitHub repository](https://github.com/UCD-SERG/rpt)
- Read the [Quarto documentation](https://quarto.org/)

## References

::: {#refs}
:::
