---
title: "Using ettbc: Emulating a Target Trial for Breast Cancer Screening"
author: "Douglas Ezra Morrison"
date: "`r Sys.Date()`"
format:
  revealjs:
    output-file: using-ettbc-revealjs.html
  html: default
  docx: default
bibliography: ../references.bib
vignette: >
  %\VignetteIndexEntry{Using ettbc: Emulating a Target Trial for Breast Cancer Screening}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(ettbc)
```

## Background

This article demonstrates how to use the `{ettbc}` package to emulate a
target trial for breast cancer screening using the **clone-censor-reweight**
methodology. The methods are based on the study by @garcia-albeniz2020, who
estimated the effect of continuing annual screening mammography on breast
cancer mortality in Medicare beneficiaries aged 70–84 years.

### The Target Trial

The study aimed to answer: *Does continuing annual screening mammography
(compared with stopping) reduce breast cancer mortality in older women?*

Because a randomised trial of this question is unlikely to be conducted, the
target trial was emulated using Medicare claims data. The key methodological
challenge is **confounding by indication**: women who continue screening tend
to be healthier and more health-conscious than those who stop.

### The Clone-Censor-Reweight Approach

The clone-censor-reweight method handles time-varying treatment and censoring
in three steps [@hernan2016]:

1. **Clone**: Each participant is replicated into two copies — one assigned
   to the STOPBASE arm (stop screening at study entry) and one to the
   CONTINUE arm (continue annual screening).

2. **Censor**: Clones are censored in their assigned arm when they deviate
   from the protocol:
   - **STOPBASE**: censored at the first screening mammogram received after
     the initial grace period.
   - **CONTINUE**: censored when more than 14 months pass without any
     mammogram.

3. **Reweight**: Inverse probability weights (IPW) are constructed to
   account for the informative censoring introduced in step 2.

The `{ettbc}` package implements the first two steps with `clone_censor()`
and provides a helper function `expand_to_long()` to prepare the dataset for
the IPW outcome analysis.

## Example Data

The package ships with three synthetic datasets that mimic the structure of
the original study data (but contain entirely simulated values).

```{r}
# One row per participant
head(cohort)
```

```{r}
# One row per screening mammogram event
head(screening_mammograms)
```

```{r}
# One row per diagnostic mammogram event
head(diagnostic_mammograms)
```

Months are numbered consecutively from 1 (January 2000) to 108
(December 2008). Participants enter the study between months 1 and 60.

```{r}
#| label: fig-entry-dist
#| fig-cap: "Distribution of study entry months in the simulated cohort."
hist(
  cohort$start_month,
  breaks = 20,
  main = "Study Entry Month",
  xlab = "Month (1 = January 2000)",
  col = "steelblue",
  border = "white"
)
```

## Step 1: Clone and Censor

`clone_censor()` takes the cohort and mammography event datasets and returns
a data frame with **two rows per participant** — one for each arm.

```{r}
cloned <- clone_censor(
  cohort,
  screening_mammograms,
  diagnostic_mammograms
)

nrow(cloned) # should be 2 × nrow(cohort)
head(cloned[, c("id", "arm", "start_month", "end_month",
                "censor_month", "fup", "died", "bc_died")])
```

The key new columns are:

- `arm`: trial arm (`"STOPBASE"` or `"CONTINUE"`)
- `end_month`: final observed month (minimum of death, administrative
  censoring, and protocol censoring)
- `censor_month`: month of protocol deviation censoring (`NA` if not censored
  for protocol non-adherence)
- `fup`: follow-up time in months
- `died`: overall mortality indicator (1 if death caused end of follow-up)
- `bc_died`: breast cancer mortality indicator

### Censoring Patterns by Arm

Because participants in STOPBASE are censored when they get a screening
mammogram, annual screeners are censored early in that arm. Conversely,
participants in CONTINUE are censored when they go too long without a
mammogram, so non-adherers are censored early.

```{r}
#| label: fig-censor-by-arm
#| fig-cap: "Follow-up time by arm and censoring type."
par(mfrow = c(1, 2))

stopbase <- cloned[cloned$arm == "STOPBASE", ]
continue <- cloned[cloned$arm == "CONTINUE", ]

hist(stopbase$fup,
  breaks = 20, main = "STOPBASE",
  xlab = "Follow-up (months)", col = "#E05C4B", border = "white",
  xlim = c(0, 110)
)
hist(continue$fup,
  breaks = 20, main = "CONTINUE",
  xlab = "Follow-up (months)", col = "#4B8FE0", border = "white",
  xlim = c(0, 110)
)
par(mfrow = c(1, 1))
```

```{r}
# Percentage censored for protocol non-adherence in each arm
prop_censored <- tapply(
  !is.na(cloned$censor_month),
  cloned$arm,
  mean
)
round(prop_censored * 100, 1)
```

As expected, most STOPBASE clones are censored (those who continued getting
annual mammograms), while only a minority of CONTINUE clones are censored
(those who stopped getting mammograms).

## Step 2: Expand to Long Format

`expand_to_long()` converts the cloned dataset to **one row per
participant-arm-month**. This format is required for the discrete-time
survival models used to estimate the outcome.

```{r}
long_data <- expand_to_long(cloned)

nrow(long_data) # total person-arm-months
head(long_data)
```

The key columns in the long dataset are:

- `month`: calendar month (1–108)
- `month2`: months since study entry (0-indexed)
- `dead_t1`: death in the next interval: `1` = yes, `0` = no, `NA` = censored
- `bc_dead_t1`: breast cancer death in the next interval
- `bc_long`: breast cancer diagnosis flag at this month

```{r}
# Total events in the long dataset
table(long_data$dead_t1, useNA = "ifany")
```

### Verifying the Long Format

A simple check: the number of observed deaths in the long dataset should
equal the number of `died == 1` rows in the cloned dataset.

```{r}
# Deaths in cloned dataset
sum(cloned$died)

# Rows with dead_t1 = 1 in long dataset
sum(long_data$dead_t1 == 1L, na.rm = TRUE)
```

## Step 3: Descriptive Analysis

Before fitting weighted outcome models, it is useful to describe the data
and examine crude survival by arm.

We first define a small helper that summarises one arm's long dataset into a
crude cumulative mortality curve. For each month since study entry it counts
the deaths, then accumulates them and divides by the number of clones at risk
at entry.

```{r}
crude_cumulative_mortality <- function(arm_long) {
  months <- 0:max(arm_long$month2)

  deaths <- vapply(
    months,
    function(t) sum(arm_long$dead_t1[arm_long$month2 == t] == 1L, na.rm = TRUE),
    numeric(1)
  )
  n_risk <- vapply(
    months,
    function(t) sum(arm_long$month2 == t),
    numeric(1)
  )

  data.frame(
    month2 = months,
    cum_mort = cumsum(deaths) / max(n_risk, 1L)
  )
}
```

With the computation factored out, plotting the two arms is a short loop over
the helper.

```{r}
#| label: fig-cumhaz
#| fig-cap: >
#|   Empirical cumulative mortality by arm (crude, without IPW).
#|   Under the null, the two arms should be similar because clones are derived
#|   from the same participants.

arms <- c("STOPBASE", "CONTINUE")
cols <- c("#E05C4B", "#4B8FE0")

plot(NA,
  xlim = c(0, 70), ylim = c(0, 0.15),
  xlab = "Months since study entry",
  ylab = "Cumulative mortality (crude)",
  main = "Crude cumulative mortality by arm"
)

for (j in seq_along(arms)) {
  curve <- crude_cumulative_mortality(long_data[long_data$arm == arms[j], ])
  lines(curve$month2, curve$cum_mort, col = cols[j], lwd = 2)
}

legend("topleft",
  legend = arms, col = cols, lwd = 2, bty = "n"
)
```

::: {.callout-note}
The crude curves above do not account for the informative censoring introduced
by the clone-censor step. The rest of this article applies **inverse
probability weights** (IPW) to correct for the different censoring mechanisms
in each arm, then fits the weighted outcome model and bootstraps a confidence
interval. See @garcia-albeniz2020 for the full analysis on the real cohort.
:::

## A Larger Simulated Cohort

The 100-participant dataset above is convenient for showing the data
structure, but it is too small to fit a stable weighted outcome model: with so
few deaths the screening-propensity model separates and the resulting weights
are erratic. For the weighted analysis we simulate a larger cohort with
`simulate_screening_cohort()`, which returns the same three linked tables.

The simulation builds in **no** screening effect on mortality, so the honest
expectation is a null result: the two arms should look alike, and the weighted
contrast should sit near no difference. What the rest of the article
demonstrates is that the machinery runs end to end and recovers that null, not
a substantive estimate; a real reproduction needs the Medicare cohort and the
paper's full covariate set.

```{r}
sim <- simulate_screening_cohort(n = 1000, max_month = 108, seed = 7)

cloned_l <- clone_censor(
  sim$cohort, sim$screening_mammograms, sim$diagnostic_mammograms
)
long_l <- expand_to_long(cloned_l)
long_l <- augment_long_covariates(
  long_l, sim$screening_mammograms, sim$diagnostic_mammograms
)

nrow(long_l) # person-arm-months
```

## Step 4: Inverse Probability Weights

`fit_screening_propensity()` fits the pooled logistic model for the
probability of a screening mammogram at each eligible month (the SAS `cann17b`
denominator), and `compute_ipw_weights()` turns those probabilities into
stabilized per-arm weights, truncated at the 99th percentile within each arm.

```{r}
propensity <- fit_screening_propensity(long_l)
weighted <- compute_ipw_weights(propensity$data, pred_prob_col = "p_scrmammo")

# Stabilized, truncated weights are centered near 1 in both arms
round(t(sapply(split(weighted$wp99, weighted$arm), summary)), 2)
```

## Step 5: Weighted Outcome Model

`fit_outcome_hr()` fits the IP-weighted pooled logistic outcome model, with
follow-up time entered through a restricted cubic spline and an arm-by-time
interaction. The reported odds ratio approximates the hazard ratio for the
STOPBASE arm relative to CONTINUE.

```{r}
hr <- fit_outcome_hr(
  weighted,
  weight_col = "wp99",
  outcome_col = "dead_t1"
)

round(c(OR = hr$or, hr$or_ci), 3)
```

As expected under a simulation with no built-in effect, the odds ratio sits
near 1 and its confidence interval covers the null.

## Step 6: Standardized Survival Curves

`predict_survival_ipw()` applies g-computation to the weighted model, averaging
the predicted survival over the cohort to produce a marginal survival curve for
each arm.

```{r}
#| label: fig-ipw-survival
#| fig-cap: >
#|   IP-weighted marginal survival by arm. The arms track each other closely,
#|   consistent with the null effect built into the simulation.

surv <- predict_survival_ipw(
  weighted,
  weight_col = "wp99",
  outcome_col = "dead_t1"
)

plot(NA,
  xlim = c(0, 72), ylim = c(0.8, 1),
  xlab = "Months since study entry",
  ylab = "Survival probability",
  main = "IP-weighted survival by arm"
)
lines(surv$month, surv$s_continue, col = "#4B8FE0", lwd = 2)
lines(surv$month, surv$s_stopbase, col = "#E05C4B", lwd = 2)
legend("bottomleft",
  legend = c("CONTINUE", "STOPBASE"),
  col = c("#4B8FE0", "#E05C4B"), lwd = 2, bty = "n"
)
```

## Step 7: Bootstrap Confidence Intervals

`bootstrap_ci()` resamples participants, recomputes the weights and survival
curves on each resample, and returns percentile confidence intervals for the
month-by-month survival difference between arms. The number of resamples is
kept small here so the article renders quickly; a real analysis uses several
hundred.

```{r}
#| label: fig-boot-diff
#| cache: true
#| fig-cap: >
#|   Bootstrap survival difference (CONTINUE minus STOPBASE) with a 95 percent
#|   percentile confidence band. The band covers zero throughout, the expected
#|   null.

boot <- bootstrap_ci(
  propensity$data,
  pred_prob_col = "p_scrmammo",
  outcome_col = "dead_t1",
  max_month = 72,
  n_boot = 50,
  seed = 1
)

plot(NA,
  xlim = c(0, 72), ylim = range(c(boot$diff_lo, boot$diff_hi), na.rm = TRUE),
  xlab = "Months since study entry",
  ylab = "Survival difference (CONTINUE - STOPBASE)",
  main = "Bootstrap survival difference with 95% CI"
)
abline(h = 0, col = "grey60", lty = 2)
polygon(
  c(boot$month, rev(boot$month)),
  c(boot$diff_lo, rev(boot$diff_hi)),
  col = "#4B8FE022", border = NA
)
lines(boot$month, boot$diff, col = "#222222", lwd = 2)
```

## Summary

The `{ettbc}` package implements the full clone-censor-reweight pipeline:

| Function | Input | Output |
|---|---|---|
| `simulate_screening_cohort()` | cohort size | three linked synthetic tables |
| `clone_censor()` | cohort + mammogram events | two rows per participant |
| `expand_to_long()` | cloned dataset | one row per participant-arm-month |
| `augment_long_covariates()` | long dataset + events | time-varying screening covariates |
| `fit_screening_propensity()` | augmented long dataset | screening probabilities |
| `compute_ipw_weights()` | long dataset + probabilities | stabilized IPW weights |
| `fit_outcome_hr()` | weighted long dataset | arm hazard ratio |
| `predict_survival_ipw()` | weighted long dataset | marginal survival curves |
| `bootstrap_ci()` | long dataset + probabilities | survival difference with CI |

: Core ettbc functions {#tbl-functions}

Run on the synthetic cohort, the pipeline recovers the null effect the
simulation was built with. Pointed at a real Medicare cohort with the paper's
full covariate set, the same functions reproduce the substantive analysis of
@garcia-albeniz2020.

The `{ettbc}` package provides two core functions for the clone-censor step
of target trial emulation:

| Function | Input | Output |
|---|---|---|
| `clone_censor()` | cohort + mammogram events | two rows per participant |
| `expand_to_long()` | cloned dataset | one row per participant-arm-month |

: Core ettbc functions {#tbl-functions}

After running these steps, the long dataset can be used with standard
weighted logistic or pooled logistic regression to estimate the
**intention-to-treat** effect of each screening strategy.

## References

::: {#refs}
:::
