---
title: "Using ettbc: Emulating a Target Trial for Breast Cancer Screening"
author: "Douglas Ezra Morrison"
date: "`r Sys.Date()`"
format:
  revealjs:
    output-file: using-ettbc-revealjs.html
  html: default
  docx: default
bibliography: ../references.bib
---

```{r setup}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
library(ettbc)
```

## Background

This article demonstrates how to use the `{ettbc}` package to emulate a
target trial for breast cancer screening using the **clone-censor-reweight**
methodology. The methods are based on the study by @garcia-albeniz2020, who
estimated the effect of continuing annual screening mammography on breast
cancer mortality in Medicare beneficiaries aged 70–84 years.

### The Target Trial

The study aimed to answer: *Does continuing annual screening mammography
(compared with stopping) reduce breast cancer mortality in older women?*

Because a randomised trial of this question is unlikely to be conducted, the
target trial was emulated using Medicare claims data. The key methodological
challenge is **confounding by indication**: women who continue screening tend
to be healthier and more health-conscious than those who stop.

### The Clone-Censor-Reweight Approach

The clone-censor-reweight method handles time-varying treatment and censoring
in three steps [@hernan2016]:

1. **Clone**: Each participant is replicated into two copies — one assigned
   to the STOPBASE arm (stop screening at study entry) and one to the
   CONTINUE arm (continue annual screening).

2. **Censor**: Clones are censored in their assigned arm when they deviate
   from the protocol:
   - **STOPBASE**: censored at the first screening mammogram received after
     the initial grace period.
   - **CONTINUE**: censored when more than 14 months pass without any
     mammogram.

3. **Reweight**: Inverse probability weights (IPW) are constructed to
   account for the informative censoring introduced in step 2.

The `{ettbc}` package implements the first two steps with `clone_censor()`
and provides a helper function `expand_to_long()` to prepare the dataset for
the IPW outcome analysis.

## Example Data

The package ships with three synthetic datasets that mimic the structure of
the original study data (but contain entirely simulated values).

```{r}
# One row per participant
head(cohort)
```

```{r}
# One row per screening mammogram event
head(screening_mammograms)
```

```{r}
# One row per diagnostic mammogram event
head(diagnostic_mammograms)
```

Months are numbered consecutively from 1 (January 2000) to 108
(December 2008). Participants enter the study between months 1 and 60.

```{r}
#| label: fig-entry-dist
#| fig-cap: "Distribution of study entry months in the simulated cohort."
hist(
  cohort$start_month,
  breaks = 20,
  main = "Study Entry Month",
  xlab = "Month (1 = January 2000)",
  col = "steelblue",
  border = "white"
)
```

## Step 1: Clone and Censor

`clone_censor()` takes the cohort and mammography event datasets and returns
a data frame with **two rows per participant** — one for each arm.

```{r}
cloned <- clone_censor(
  cohort,
  screening_mammograms,
  diagnostic_mammograms
)

nrow(cloned) # should be 2 × nrow(cohort)
head(cloned[, c("id", "arm", "start_month", "end_month",
                "censor_month", "fup", "died", "bc_died")])
```

The key new columns are:

- `arm`: trial arm (`"STOPBASE"` or `"CONTINUE"`)
- `end_month`: final observed month (minimum of death, administrative
  censoring, and protocol censoring)
- `censor_month`: month of protocol deviation censoring (`NA` if not censored
  for protocol non-adherence)
- `fup`: follow-up time in months
- `died`: overall mortality indicator (1 if death caused end of follow-up)
- `bc_died`: breast cancer mortality indicator

### Censoring Patterns by Arm

Because participants in STOPBASE are censored when they get a screening
mammogram, annual screeners are censored early in that arm. Conversely,
participants in CONTINUE are censored when they go too long without a
mammogram, so non-adherers are censored early.

```{r}
#| label: fig-censor-by-arm
#| fig-cap: "Follow-up time by arm and censoring type."
par(mfrow = c(1, 2))

stopbase <- cloned[cloned$arm == "STOPBASE", ]
continue <- cloned[cloned$arm == "CONTINUE", ]

hist(stopbase$fup,
  breaks = 20, main = "STOPBASE",
  xlab = "Follow-up (months)", col = "#E05C4B", border = "white",
  xlim = c(0, 110)
)
hist(continue$fup,
  breaks = 20, main = "CONTINUE",
  xlab = "Follow-up (months)", col = "#4B8FE0", border = "white",
  xlim = c(0, 110)
)
par(mfrow = c(1, 1))
```

```{r}
# Percentage censored for protocol non-adherence in each arm
prop_censored <- tapply(
  !is.na(cloned$censor_month),
  cloned$arm,
  mean
)
round(prop_censored * 100, 1)
```

As expected, most STOPBASE clones are censored (those who continued getting
annual mammograms), while only a minority of CONTINUE clones are censored
(those who stopped getting mammograms).

## Step 2: Expand to Long Format

`expand_to_long()` converts the cloned dataset to **one row per
participant-arm-month**. This format is required for the discrete-time
survival models used to estimate the outcome.

```{r}
long_data <- expand_to_long(cloned)

nrow(long_data) # total person-arm-months
head(long_data)
```

The key columns in the long dataset are:

- `month`: calendar month (1–108)
- `month2`: months since study entry (0-indexed)
- `dead_t1`: death in the next interval: `1` = yes, `0` = no, `NA` = censored
- `bc_dead_t1`: breast cancer death in the next interval
- `bc_long`: breast cancer diagnosis flag at this month

```{r}
# Total events in the long dataset
table(long_data$dead_t1, useNA = "ifany")
```

### Verifying the Long Format

A simple check: the number of observed deaths in the long dataset should
equal the number of `died == 1` rows in the cloned dataset.

```{r}
# Deaths in cloned dataset
sum(cloned$died)

# Rows with dead_t1 = 1 in long dataset
sum(long_data$dead_t1 == 1L, na.rm = TRUE)
```

## Step 3: Descriptive Analysis

Before fitting weighted outcome models, it is useful to describe the data
and examine crude survival by arm.

```{r}
#| label: fig-cumhaz
#| fig-cap: >
#|   Empirical cumulative mortality by arm (crude, without IPW).
#|   Under the null, the two arms should be similar because clones are derived
#|   from the same participants.

# Compute monthly event rate per arm
arms <- c("STOPBASE", "CONTINUE")
cols <- c("#E05C4B", "#4B8FE0")

plot(NA,
  xlim = c(0, 70), ylim = c(0, 0.15),
  xlab = "Months since study entry",
  ylab = "Cumulative mortality (crude)",
  main = "Crude cumulative mortality by arm"
)

for (j in seq_along(arms)) {
  sub <- long_data[long_data$arm == arms[j], ]
  # Month-by-month event counts
  max_t <- max(sub$month2)
  cum_mort <- numeric(max_t + 1L)
  n_risk <- numeric(max_t + 1L)

  for (t in 0:max_t) {
    rows_t <- sub[sub$month2 == t, ]
    n_risk[t + 1L] <- nrow(rows_t)
    cum_mort[t + 1L] <- sum(rows_t$dead_t1 == 1L, na.rm = TRUE)
  }
  cum_mort <- cumsum(cum_mort) / max(n_risk, 1L)
  lines(0:max_t, cum_mort, col = cols[j], lwd = 2)
}

legend("topleft",
  legend = arms, col = cols, lwd = 2, bty = "n"
)
```

::: {.callout-note}
The crude curves above do not account for the informative censoring introduced
by the clone-censor step. In a complete analysis, **inverse probability
weights** (IPW) would be applied to correct for the different censoring
mechanisms in each arm. See @garcia-albeniz2020 for the full
weighted analysis.
:::

## Summary

The `{ettbc}` package provides two core functions for the clone-censor step
of target trial emulation:

| Function | Input | Output |
|---|---|---|
| `clone_censor()` | cohort + mammogram events | two rows per participant |
| `expand_to_long()` | cloned dataset | one row per participant-arm-month |

: Core ettbc functions {#tbl-functions}

After running these steps, the long dataset can be used with standard
weighted logistic or pooled logistic regression to estimate the
**intention-to-treat** effect of each screening strategy.

## References

::: {#refs}
:::
